Pop-up cycleways provide a peak into post COVID-19 transport futures

### 2020-05-12 conversation.Rmd
Update conversation

---
title: "Pop-up cycleways provide a peak into post COVID-19 transport futures"
---

As nations prepare to emerge from months of COVID-19 induced lockdowns, it is becoming clear life cannot simply return to normal. The pandemic has transformed all sectors of the economy, not least transport.

In the coming weeks and months, walking and cycling will play a vital role in enabling the reopening of many countries, as continued social distancing measures will mean that public transport options are likely to run at a fraction of their intended capacity. There is a window of opportunity here that could be missed if roads become busy and dangerous again before new active transport habits become embedded.

Local and national governments are scrambling to act. In the UK, £2 billion investment in measures to promote walking and cycling has been announced. £250 million of this is for "emergency interventions to make cycling and walking safer" as [Transport Secretary Grant Shapps put it](https://www.gov.uk/government/speeches/transport-secretarys-statement-on-coronavirus-covid-19-9-may-2020). New [guidance from the Department for Transport](https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19) advocates "measures to reallocate road space to people walking and cycling".

@@

Local authorities urgently need answers on how this funding should be spent, and where road space reallocation should be prioritised. New cyclinginfrastructure which is developed based on rigorous analysis of the best available data, in combination with local knowledge and community engagement, are more likely to be effective than plans based on intuition alone. Recognising the importance of a strong and actionable evidence-base, and the window of opportunity, we used our experience developing tools that are already being used by local authorities, notably the multi-institution PCT and the CyIPT, to develop methods and outputs taylored for the new transport planning landscape.

<image id="334100" align="centre" caption="Screenshot from the website salfordliveablestreets.commonplace.is to support local responses to the COVID-19 pandemic, including the prioritisation of pop-up active transport infrastructure." />

In a pre-print [paper](https://osf.io/7wjb6) on methods to prioritise pop-up road space reallocation schemes, we have classified pop-up interventions into three categories. Firstly, in streets that are wide or have multiple lanes, roadspace can be reallocated to widen pavements or introduce pop-up cycleways, such as outside a hospital in [Leicester](https://www.leicestermercury.co.uk/news/leicester-news/light-lockdown-traffic-leicester-allows-4083625). Secondly, physical barriers can be used to make routes for walking and cycling [faster and more direct](https://cyclehighways.eu/design-and-build/infrastructure/filtered-permeability.html) than the routes for motor vehicles. This can be seen in Tower Hamlets, London, where simple barriers like planters are used to good effect. Thirdly, roads can be closed altogether to motor vehicles, as is happening in [Deansgate, Manchester](https://www.manchestereveningnews.co.uk/news/greater-manchester-news/part-deansgate-close-cars-city-18178175).

The first of these three options is the most simple to implement,  with relatively few implications for traffic flow. We have mapped out **link** the top locations for this reallocation of road space in ten major cities across England, with the intention of rolling out these methods across the country in the near future. Having identified roads where a spare lane could be re-purposed for cycling, we have assessed the cycle potential of these streets based on census data for travel to work and cycling uptake scenarios under the UK Government’s national target of doubling cycling levels by 2025. The length of road segments is also used to filter the top schemes.

The use of open data and publicly accessible models is a key aspect of this research. Given the global nature of the challenges we face, open and collaborative research is vital so that methods can be applied by as wide a user base as possible. The potential for international application can be seen for example in [research for the World Health Organisation](https://atfutures.github.io/upthat/articles/adaptation.html) modelling cycling uptake in cities in low income countries. Within the UK, [50+ Local Authorities](https://npct.github.io/pct-shiny/regions_www/www/static/03d_other_reports/2019-use-of-pct-report.pdf) are now using the PCT to inform their strategic cycle network plans.

Results from our [methods](https://osf.io/7wjb6) show that many cities have 'spare lanes' under the current levels of traffic in city centres and near key worker destinations such as hospitals. These results could support the prioritisation of vital resources dedicated to pop-up infrastructure. With the situation rapidly developing, some of the roads shown here, such as [Kirkstall Road](https://www.transportxtra.com/publications/local-transport-today/news/65382/leeds-to-pilot-enhanced-cycling-safety-u) (in the top 10 results our methods generated for Leeds), have already been identified by Local Authorities as targets for intervention. As the recovery continues, we hope our methods will support cities to take the opportunities that the crisis presents for active travel.

<image id="334113" align="centre" caption="Roads in Leeds with at least one spare lane that could be reallocate to provide space for active travel, and locations of hospitals" />

@@

With the rush to act, there is a great need for evidence of where new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the ‘pop-up’ infrastructure that goes in is effective, safe, and placed where it is most needed. Agile research can respond to rapidly changing conditions, as government priorities shift and different sectors of the economy reopen. Measures can be undertaken on a temporary basis, but if action is prioritised in places where there is a long-term need for cycle improvements, there is a greater chance that they become permanent. In Paris, Covid-19 related measures aim to contribute towards Mayor Anne Hidalgo's long-term [Plan Velo](https://www.curbed.com/2020/1/15/21065343/bike-paris-cycling-anne-hidalgo).

>>>>

---
title: "Pop-up cycleways provide a peak into post COVID-19 transport futures"
---
As cities across the world prepare to emerge from lockdown, continued social distancing measures will mean that public transport options are likely to run at a fraction of their intended capacity. This is a crucial moment for urban cycling.

There is a window of opportunity to change urban travel. Acting fast, [before roads become busy again](https://www.forbes.com/sites/carltonreid/2020/05/07/car-traffic-rise-of-11-shows-urgent-need-for-pop-ups-to-protect-cyclists-and-pedestrians/amp/), will enable new and healthy active transport habits to become embedded long-term.

Our research at the University of Leeds is investigating how best to make rapid changes to urban cycling infrastructure. We have developed [methods](https://osf.io/7wjb6) to identify roads where there is space for pop-up infrastructure, and where it is most needed.

The UK government has announced £2 billion of investment in measures to promote walking and cycling in England over the next 5 years. £250 million has been allocated for "emergency interventions to make cycling and walking safer" as [Transport Secretary Grant Shapps put it](https://www.gov.uk/government/speeches/transport-secretarys-statement-on-coronavirus-covid-19-9-may-2020), with similar commitments being made by [Scottish](https://www.transport.gov.scot/news/10-million-to-support-pop-up-active-travel-infrastructure/) and [Welsh](https://gov.wales/councils-challenged-change-wales-transport-environment-during-coronavirus-lockdown) governments.

@@

Local authorities urgently need to decide how this funding should be spent, and where the reallocation of road space should be prioritised. New cycling infrastructure is more likely to be effective when it is developed based on analysis of the best available data, in combination with local knowledge and community engagement.

@@

If action is prioritised in places where there is a long-term need for cycle improvements, there is a greater chance that these pop-up developments can become permanent. In Paris, COVID-19 related measures aim to contribute towards Mayor Anne Hidalgo's long-term [Plan Velo](https://www.curbed.com/2020/1/15/21065343/bike-paris-cycling-anne-hidalgo).

We have experience of building tools and software to help highlight roads that may be suitable candidates for pop-up infrastructure. The [Propensity to Cycle Tool](https://www.pct.bike/), built in collaboration with other universities and funded by the Department for Transport, helps local authorities identify [where cycle networks](https://npct.github.io/pct-shiny/regions_www/www/static/03d_other_reports/2019-use-of-pct-report.pdf) can be established.

Now, we have used this knowledge to generate results that can support investment in pop-up infrastructure. As with the PCT, the visualised and interpreted using interactive maps. The accessible and open nature of the results encourages community engagement and online debate, based on the evidence.

<tweet url="https://twitter.com/robinlovelace/status/1258411769682366466" />

In streets that are wide or have multiple lanes, road space can be reallocated to widen pavements or introduce pop-up cycleways, such as outside a hospital in [Leicester](https://www.leicestermercury.co.uk/news/leicester-news/light-lockdown-traffic-leicester-allows-4083625). This is simple to implement,  with relatively few implications for traffic flow.

[Publicly available and interactive preliminary results](https://cyipt.github.io/popupCycleways/) for 10 cities highlight the potential of the methods and provide an opportunity for people with local knowledge to feedback, via ['issues' on the collaboration platform GitHub](https://github.com/cyipt/popupCycleways/issues) (requires a [GitHub](https://github.com/) account). Our work has already been [used](https://www.cyclinguk.org/press-release/100-miles-pop-cycle-lanes-could-enable-millions-cycle-safely-post-lockdown) to support cycling advocates to push pop-up infrastructure where it is most needed. An [interactive map of local interventions nationwide](https://www.cyclinguk.org/covid-19-safe-space-social-distancing) highlights what is possible and provides a basis for engaging with local authorities to encourage swift and effective action to encourage walking and cycling. Our research will contribute to the effort by strengthening the case for pop-up cycleways and highlighting available road space and locations with high long-term cycling potential.

The results are based on two main inputs: the Propensity to Cycle Tools's analysis of Census data under the UK Government’s [national target](https://www.gov.uk/government/publications/cycling-and-walking-investment-strategy) of doubling cycling levels by 2025, and data on the characteristics of road sections from the CyIPT project. Roads with spare lanes were defined as roads on which there are at least two lanes going in the same direction. The methods, based on open data, open source software and reproducible code, could scale nationwide and to other cities worldwide. Further refinements of the methods could include the prioritisation of roads to 'key destinations' such as hospitals (with hospitals in Leeds shown in the map below) and the inclusion of other relevant variables such as road width and safety statistics.

<image id="334113" align="centre" caption="Roads in Leeds with at least one spare lane that could be reallocated to provide space for active travel, and locations of hospitals" />

The use of open data and publicly accessible models is a key aspect of this research. Given the global nature of the challenges we face, open and collaborative research is vital so that methods can be applied by as wide a user base as possible. The potential for international application can be seen for example in [research for the World Health Organisation](https://atfutures.github.io/upthat/articles/adaptation.html) modelling cycling uptake in cities in low income countries. Within the UK, [over 50 Local Authorities](https://npct.github.io/pct-shiny/regions_www/www/static/03d_other_reports/2019-use-of-pct-report.pdf) are now using the PCT to inform their strategic cycle network plans.

An important [finding](https://osf.io/7wjb6) is that many cities have excess road space, under reduced traffic conditions, on routes leading to key destinations. Identifying such roads, like [Kirkstall Road](https://www.openstreetmap.org/way/7991370#map=15/53.8049/-1.5718), highlighted by the results shown in the map above and already part of plans by [Leeds City Council](https://news.leeds.gov.uk/news/new-measures-to-support-social-distancing-for-people-walking-and-cycling-in-leeds) to become a [pilot pop-up cycleway](https://www.transportxtra.com/publications/local-transport-today/news/65382/leeds-to-pilot-enhanced-cycling-safety-u), could be the key to effective investment in pop-up infrastructure. The evidence generated by our methods, in combination with community engagement, can ensure that transport interventions made during the crisis are of maximum benefit now and in the post pandemic world.
